Estimating Neighborhood Rents with Scraped Data

### 2018-06-04 manuscript/index.Rmd
work as of 6/4 morning

---
title: "Estimating Neighborhood Rents with Scraped Data"
author: "Chris Hess"
date: "June 2, 2018"
output: 
  html_document:
    toc: true
    toc_float: true
    includes:
      in_header: "header.html"
      after_body: "../footer.html"
abstract: "Online classified ads are a key source of information about local rental housing markets. This research asks what temporal and spatial dynamics explain neighborhood variations in one-bedroom rents in Seattle, WA, and contributes a novel approach to forecasting quarterly neighborhood rent levels using Bayesian models estimated with integrated nested Laplace approximations (INLA). All models were trained on data for 2017 Q2 through 2018 Q1, with forecasts for 2018 Q2 compared against the observed-to-date data for the current quarter (i.e. 2018 Q2). The top model assumes that spatial and temporal dynamics are separable, i.e. independent of each other, and non-interactive. This spatial random-effects model with AR(1) temporal structure achieves a mean absolute error (MAE) of about $90 for the training data and $145 for the test data."
---

@@

Demand for rental housing units in Seattle, WA surged in the past decade, due to relative scarcity in available housing amidst a period of historic population growth. Growing shares of households are unable to buy a home short of "driving till they qualify", and among households who rent a substantial share pay 30%, if not more of their incomes on their housing costs. The rising cost of housing therefore has a direct role in neighborhood processes of "demographic inversion" like urban gentrification and suburbanizing poverty. American Community Survey data have common currency among scholars and practitioners interested in metropolitan housing markets, but neverthless possess a substantial limitation for research in tight housing markets like Seattle---that is, at least two years elapse between data collection and publication, if not more. This publication lag for ACS data prohibits analyzing any market dynamics beyond long-term trends, particularly for estimates of smaller areas of aggregation like neighborhoods where ACS uses 5-year rolling-average methods as well. While more up-to-date options for neighborhood rent data exist through privately-held companies, such data can be somewhat of a black box in their creation and may be biased towards the costs of corporate-held properties. This would-be problem for academic researchers creates an opportunity for applying web scraping methods to generate a novel data on housing availability and costs.

This study aims to demonstrate why scraped rental housing listings are a valuable source of data in subject areas such as housing studies and neighborhood demography through a set of descriptive analyses and models of quarterly rent levels among census tracts in Seattle, WA. Summaries of the Craigslist data show agreement with existing sources in their distribution but do suggest higher median levels for most neighborhoods, likely given the listings' recent collection relative to other data. In a more general sense, the methods and frameworks used in this study also offer a template for efficiently collecting web data, conducting reproducible research, and publishing interactive results to GitHub. I start by briefly reviewing some background relevant to studying neighborhood rental housing. I then outline a process for automated web scraping of a rental listing source like Craigslist using a program called Helena, and compare aggregates from the collected rental listings with other data to assess rent agreement in measures. After describing the data and process, I present a set of models predicting neighborhood rent levels to understand different how different temporal and spatial structures fit the data. I conclude by discussing the promise of web data for subfields like housing studies and the value of R Markdown and GitHub for social science research.

@@

Existing public data sources like the American Community Survey or decennial census theoretically have two sources of error relative to the ``true'' median rent in a given overall market and to an even greater degree among neighborhood approximations like census tracts. Census sources of rental information cannot provide insight into any change in rental cost associated with newer developments becoming available (at least not until considerable time has already passed) given the temporal delay in data publication (i.e. ACS estimates are usually at least 2 years old, decennial many more). This feature of Census data is salient given newer housing units' generally higher rents and greater-than-average vacancy, both factors which lead to underestimating the present levels of rents for available housing units when either the ACS or decennial census are used as a current indicator. For already-leased rental housing units that are sampled, the temporal lag generates measurement error such that data on a given building's current asking rents may be underestimated relative to current tenants' payment levels.

@@

[Helena](http://helena-lang.org) is a high-level programming language for end users to automate web scraping as a method of data collection. Users write programs in an extension for Google Chrome by demonstrating the first row of data they want to collect using clicks, form interactions and highlighting values. The extension translates an initial demonstration into a starting script written in Helena, which a user can further modify or set to run on a regular schedule. Though Helena programs collect data roughly at the speed they are recorded at, users can run multiple programs in parallel to increase the volume of data collected. Data are stored on an AWS server where users can later query either the full table of scraped data or a subset scraped within a specified number of hours.

@@

The key steps of any data process can be distilled into three tasks: a.) _extract_ a table of interest from a specified origin source, b.) _transform_ it to support the desire analyses and c.) _load_ the transformed data to a storage destination for future use. In the present case, a program extracts scraped listings observed since the database 
<br>

@@

@@

![Figure 2. Time-series of median rent (all bedroom sizes) for Seattle, 1997-2018. Historical values adjusted with CPI; Craigslist estimates are quarterly, Dupre+Scott are semi-annual and ACS are annual (i.e. respective 1-Year Estimates for each year period).](../output/rent_lineplot.png)

@@

![Figure 3. Chloropleth map of neighborhood 1-bedroom rents across Craigslist, Dupre+Scott and ACS 2012-2016 5-Year Estimates. Craigslist represents median asking rent among all one-bedroom listings, Dupre+Scott consists of median contract rents among one-bedroom units in 20+ unit properties, while ACS releases median gross (asking) rents for all one-bedroom renter-occupied housing units sampled.](../output/rent_map_across_1B.png)

@@

![Figure 4. Scatter plot of neighborhood 1-bedroom rents across Craigslist and Dupre+Scott. Craigslist represents median asking rent among all one-bedroom listings, Dupre+Scott consists of median contract rents among one-bedroom units in 20+ unit properties.](../output/rent_scatter_across_1B.png)

@@

@@

`r knitr::kable(fit_stats)`

>>>>

---
title: "Estimating Neighborhood Rents with Scraped Data"
author: "Chris Hess"
date: "June 2, 2018"
output: 
  html_document:
    toc: true
    toc_float: true
    includes:
      in_header: "header.html"
      after_body: "../footer.html"
abstract: "Online classified ads are a key source of information about local rental housing markets. This research first assesses the agreement of census tract rent estimates based on scraped rental advertisements with existing data sources. The second component of the paper investivates temporal and spatial dynamics in one-bedroom rents among Seattle neighborhoods using spatial models. All models' posterior distribution are estimated with integrated nested Laplace approximations (INLA). Training data for models cover quarterly tract estimates for 2017 Q2 through 2018 Q1, while the test data are 2018 Q2 median values as observed through June 3rd. The top model assumes that spatial and temporal dynamics are separable, i.e. independent of each other, and non-interactive. This spatial random-effects model with AR(1) temporal structure achieves a mean absolute error (MAE) of about $95 for the training data and $140 for the test data."
---

@@

Demand for rental housing units in Seattle, WA surged in the past decade, due to relative scarcity in available housing amidst a period of historic population growth. Substantial appreciation for almost 2 years straight raised Seattle home prices to levels where [most households are no longer able to buy a home](https://www.seattletimes.com/business/real-estate/seattle-area-homebuyers-needed-11000-pay-raise-in-past-year-to-keep-up-with-rising-prices/?utm_source=The+Seattle+Times&utm_campaign=de86e2d581-Morning_Brief_11_28_2017&utm_medium=email&utm_term=0_5beb38b61e-de86e2d581-122931741) short of "driving till they qualify". More and more households must rent to live in Seattle, and a [almost half](http://www.jchs.harvard.edu/ARH_2017_cost_burdens_by_metro) of these renter households pay 30% of their incomes on housing costs, if not more. Beyond individual households, the rising cost of housing in Seattle and in other cities has a role in neighborhood processes of "demographic inversion" like urban gentrification and [suburbanizing poverty](https://www.brookings.edu/testimonies/the-changing-geography-of-us-poverty/) too.

[American Community Survey (ACS)](https://www.census.gov/programs-surveys/acs/) data have common currency among scholars and practitioners interested in metropolitan housing markets, but neverthless possess a substantial limitation for research in tight housing markets like Seattle---that is, at least two years elapse between data collection and publication, if not more. This publication lag for ACS data prohibits analyzing any market dynamics beyond long-term trends, particularly for estimates of smaller areas of aggregation like neighborhoods where ACS uses 5-year rolling-average methods as well. While more up-to-date options for neighborhood rent data exist through privately-held companies, such data can be somewhat of a black box in their creation and may be biased towards the costs of corporate-held properties. This would-be problem for academic researchers creates an opportunity for applying web scraping methods to generate a novel data on housing availability and costs.

This study demonstrates why scraped rental housing listings are a valuable source of data in subject areas such as housing studies and neighborhood demography. I draw on a set of descriptive analyses and models of quarterly rent levels among census tracts in Seattle, WA for this research objective. Summaries of the Craigslist data show agreement with existing sources in their distribution but do suggest higher median levels for most neighborhoods, likely given the listings' recent collection relative to other data. The smoothing models use the panel of quarterly neighborhood estimates' spatial and temporal structure to understand the underlying generative process, finding that a combination of local smoothing and autoregression (order 1) predict future levels with an average error of about $140 or 8%.

In a more general sense, the methods and frameworks used in this study also offer a template for efficiently collecting web data, conducting reproducible research, and publishing interactive results to GitHub. I start by briefly reviewing some background relevant to studying neighborhood rental housing. I then outline a process for automated web scraping of a rental listing source like Craigslist using a program called Helena, and compare aggregates from the collected rental listings with other data to assess rent agreement in measures. After describing the data and process, I present a set of models predicting neighborhood rent levels to understand different how different temporal and spatial structures fit the data. I conclude by discussing the promise of web data for subfields like housing studies and the value of R Markdown and GitHub for social science research.

@@

Existing public data are essentially limited to the American Community Survey (ACS) following the cessation of long-form decennial censuses in 2000. ACS data theoretically have two sources of error relative to the ``true'' median rent in a given overall market and to an even greater degree among neighborhood approximations like census tracts. Census sources of rental information cannot provide insight into any change in rental cost associated with newer developments becoming available (at least not until considerable time has already passed) given the temporal delay in data publication (i.e. ACS estimates are usually at least 2 years old, decennial many more). This feature of Census data is salient given newer housing units' generally higher rents and greater-than-average vacancy, both factors which lead to underestimating the present levels of rents for available housing units when either the ACS or decennial census are used as a current indicator. For already-leased rental housing units that are sampled, the temporal lag generates measurement error such that data on a given building's current asking rents may be underestimated relative to current tenants' payment levels.

@@

[Helena](http://helena-lang.org) is a high-level programming language for end users to automate web scraping as a method of data collection. Users write programs in an extension for Google Chrome by demonstrating the first row of data they want to collect using clicks, form interactions and highlighting values. The extension translates an initial demonstration into a starting script written in Helena, which a user can further modify or set to run on a regular schedule. Though Helena programs collect data roughly at the speed they are recorded at, users can run multiple programs in parallel to increase the volume of data collected. Data accumulate on a remote server (soon-to-be AWS) where users can later query either the full table of scraped data or a subset scraped within a specified number of hours.

@@

The key steps of any data process can be distilled into three tasks: a.) _extract_ a table of interest from a specified origin source, b.) _transform_ it to support the desire analyses and c.) _load_ the transformed data to a storage destination for future use.

In the present case, a program extracts scraped listings observed since the data updated by sending a download request to the scraper's accumulated data in the form of a URL with a suffix indicating the number of hours that passed since the last processing. The processing routine then transforms the data by cleaning up some of the scraped fields and appending spatial information through a multi-step geocoding proces. If address information was successfully scraped from the listing metadata, the script makes a call to Google Maps Geocoding API using this listing feature and a constraint to search within Washington state. When this first step is not successful at geocoding to a specific address or cross-streets, or where there is no address in the listing metadata, we use SmartyStreet's Address Extraction API on the listing's full scraped text to see if an address was mentioned there. We are able to geocode 95\% of listings to some degree of precision, with the vast majority precise at least to Zip9 geography. All listings used in the following analyses are geocoded with precision at least to Zip9 geography. After geocoding the data and ensuring there are no duplicates among the table of new listings, the script loads the processed listings into the existing database.

@@

```{r n_heatmap, echo=FALSE, message=FALSE, warning=FALSE}
#determine config (dev or github)
if(file.exists("../../data/cl/craigslistDB.sqlite")){
  source("../helpers/n_heatmap.R") #if config, refresh plot pdf
}
```

![Figure 2. Heatmap of N Seattle Craigslist one-bedroom listings scraped per day time-series.](../output/n_heatmap.png)

@@

![Figure 3. Time-series of median rent (all bedroom sizes) for Seattle, 1997-2018. Historical values adjusted with CPI; Craigslist estimates are quarterly, Dupre+Scott are semi-annual and ACS are annual (i.e. respective 1-Year Estimates for each year period).](../output/rent_lineplot.png)

@@

![Figure 4. Chloropleth map of neighborhood 1-bedroom rents across Craigslist, Dupre+Scott and ACS 2012-2016 5-Year Estimates. Craigslist represents median asking rent among all one-bedroom listings, Dupre+Scott consists of median contract rents among one-bedroom units in 20+ unit properties, while ACS releases median gross (asking) rents for all one-bedroom renter-occupied housing units sampled.](../output/rent_map_across_1B.png)

@@

![Figure 5. Scatter plot of neighborhood 1-bedroom rents across Craigslist and Dupre+Scott. Craigslist represents median asking rent among all one-bedroom listings, Dupre+Scott consists of median contract rents among one-bedroom units in 20+ unit properties.](../output/rent_scatter_across_1B.png)

@@

- Need to build tables from output
- Need to do write up for tables
- Need to add graphic for predicted values

@@

`r knitr::kable(fit_stats %>% select(ends_with("rmse")))`

`r knitr::kable(fit_stats %>% select(ends_with("mae")))`
